Assessing Power of the California Microhap Baseline for GSI (and calculating Fst)

### 2024-05-14 002-GSI_and_Fst.Rmd
Big overhaul just before I gave lab meeting about this

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    #final_baseline = "data/cali-chinook-baseline.rds",
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf"
  )
}

@@

no_lfar_ass_table <- plot_self_assignment_table(top_ass, pop_labels, collection_order)

no_lfar_ass_table_80 <- top_ass %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

no_lfar_ass_table_Rosa <- top_ass %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

And then do the same for the data with the LFAR:
```{r}
lfar_sa <- self_assign(base_with_lfar, gen_start_col = 5)
lfar_top_ass <- get_top_assignments(lfar_sa, collection_order)
lfar_ass_table <- plot_self_assignment_table(lfar_top_ass, pop_labels, collection_order)

lfar_ass_table_80 <- lfar_top_ass %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

lfar_ass_table_Rosa <- lfar_top_ass %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

And, once again, we want to investigate the ones that are in the wrong repunit, and look at their
scaled_repu_likelihoods, the others.
```{r}
source("R/colors.R")
ta_corr <- top_ass %>%
  filter(repunit == inferred_repunit)
ta_wrong <- top_ass %>%
  filter(repunit != inferred_repunit) %>%
  arrange(collection, repu_sclike) %>%
  group_by(collection) %>%
  mutate(y = 190 + (70/9) * 1:n())

ggplot() +
  geom_histogram(
    data = ta_corr, 
    mapping = aes(x = repu_sclike, fill = inferred_repunit)
  ) +
  geom_point(
    data = ta_wrong,
    mapping = aes(x = repu_sclike, y = y, fill = inferred_repunit),
    size = 3, 
    shape = 21
  ) +
  facet_wrap(~coll_f, ncol = 3) +
  scale_fill_manual(values = repunit_colors)
```
Which looks to me like there are really only 7 here that we are not getting right.
Those others we will filter out and make a table of them:

```{r}
tafilt <- top_ass %>%
  filter(!(repunit != inferred_repunit & repu_sclike > 0.99))
```

Here we calculate the number of correctly and incorrectly assigned fish:
```{r}
tafilt %>%
  mutate(correct = repunit == inferred_repunit) %>%
  count(correct) %>%
  mutate(fract = n / sum(n))
```

And now we will make a table of those:
```{r}
# get the function
source("R/taf-prep.R")
source("R/table-as-figure.R")

# get the RC_groups
RC_groups <- pop_labels %>%
  select(-old_name) %>%
  distinct() %>%
  rename(external_group = run_timing_group, internal_group = repunit, label = collection) %>%
  select(external_group, internal_group, label)

X2 <- tafilt %>%
  mutate(
    collection_f = factor(collection, levels = collection_order),
    inferred_coll_f = factor(inferred_collection, levels = collection_order)
  ) %>%
  count(collection_f, inferred_coll_f, .drop = FALSE) %>%
  taf_prep()

Rubias_for_later <- X2 # keeping this for later...

# get the result
TAF2 <- table_as_figure(
  X = X2, 
  RC_groups = RC_groups,
  external_colors = run_time_colors,
  internal_colors = repunit_colors, 
  plot_margins = c(1, 0.1, 0.7, 0.1)
)

# and plot it
g <- plot_grid(
  TAF2$full_plot,
  plot_grid(
    get_legend(TAF2$for_external_legend),
    get_legend(TAF2$for_internal_legend),
    nrow = 1
  ),
  nrow = 2,
  rel_heights = c(7,3)
)

g
```

That is a considerable improvement. Let's write it out:
```{r}
ggsave(g, filename = output_list$assig_table, width = 8, height = 10)
```

# Now, we repeat that while adding the LFAR markers in there

Then, all the step from above:
```{r}
# Do self-assignment:
full_plus_lfar_sa <- self_assign(base_with_lfar, gen_start_col = 5)

repu_likes <- full_plus_lfar_sa %>%
  group_by(indiv, collection, repunit, inferred_repunit) %>%
  summarise(repu_sclike = sum(scaled_likelihood)) %>%
  ungroup()

top_ass_w_lfar <- full_plus_lfar_sa %>%
  group_by(indiv, collection, repunit, inferred_repunit) %>%
  mutate(repu_sclike = sum(scaled_likelihood)) %>%
  group_by(indiv) %>%
  filter(repu_sclike == max(repu_sclike)) %>%
  arrange(indiv, desc(scaled_likelihood)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(coll_f = factor(collection, levels = collection_order))

@@

We start by tallying up the misassigned fish---those that have a repunit and an
inferred repunit that differ:
```{r}
misassigned_ones <- top_ass_w_lfar %>%
  filter(repunit != inferred_repunit) %>%
  count(repunit, inferred_repunit)

@@

tmp <- lapply(1:nrow(MO2), function(i) {
  repu_likes %>%
    filter(
      (repunit == MO2$Foc_repunit[i] | repunit == MO2$Foc_inferred_repunit[i]) &
        (inferred_repunit == MO2$Foc_repunit[i] | inferred_repunit == MO2$Foc_inferred_repunit[i])
    ) %>%
    mutate(
      Foc_repunit =  MO2$Foc_repunit[i],
      Foc_inferred_repunit = MO2$Foc_inferred_repunit[i],
      .before = repunit
    )
}) %>%
  bind_rows() %>%
  arrange(indiv, Foc_repunit, Foc_inferred_repunit)

for_plots <- tmp %>%
  mutate(
    rep = ifelse(Foc_repunit == repunit, "rep", "inf"),
    inf = ifelse(Foc_inferred_repunit == inferred_repunit, "inf", "rep")
  ) %>%
  #count(Foc_repunit, Foc_inferred_repunit, repunit, inferred_repunit, rep, inf)
  select(-repunit, -inferred_repunit) %>%
  pivot_wider(names_from = c(rep, inf), names_sep = "_to_", values_from = repu_sclike) %>%
  mutate(condition = str_c(Foc_repunit, " misassigned to ", Foc_inferred_repunit), .after = Foc_inferred_repunit)
```

@@

In order to do this we need to convert the data to a lumped format.
This means making integers of the alleles and then catenating them.
One locus has 11 allele, so we should use two digits for each allele.  We drop the
LFAR from this, because those were chosen specifically to be weird in the late fall.
```{r}
full_base_long <- full_base2 %>% 
  select(-starts_with("NC_037130.1")) %>%  # drop the LFARs
  mutate(
    Pop = as.integer(factor(collection, levels = collection_order)),
    .before = collection
  ) %>%
  select(-repunit, -collection, -sample_type) %>%
  pivot_longer(
    cols = c(-indiv, -Pop),
    names_to = c("locus", "gene_copy"),
    names_pattern = "^(.*)_([12])$"
  )

@@

# long format with no missing data
long_genos <- full_base2 %>%
select(-repunit, -sample_type) %>%
pivot_longer(
cols = -c(indiv, collection),
names_to = c("locus", "gene_copy"),
values_to = "allele",
names_pattern = "(.*)[_.]([12])$"
) %>%
filter(!is.na(allele))

@@

# we need to get colors and things for the fill.  We just take the
# rubias output that was run through taf_prep and we keep the lower
# diagonal and we set all the cell_labels to NA.
lower_tri_base <- Rubias_for_later %>%
filter(as.integer(col_label) <= as.integer(row_label)) %>%
mutate(cell_label = NA_character_)

@@

@@

```{r}
       gsi_fst_fig_for_paper <- plot_grid(
TAF2$full_plot,
plot_grid(
get_legend(TAF2$for_external_legend),
get_legend(TAF2$for_internal_legend),
nrow = 1
),
fst_with_scatters,
nrow = 3,
rel_heights = c(6,1.5,6)
)

>>>>

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    #final_baseline = "data/cali-chinook-baseline.rds",
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    ass_table80 = "results/GSI_and_Fst/ass-table-80.pdf",
    rosa_gsi_table = "results/GSI_and_Fst/rosa-gsi-table.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf",
    ass_table80_crop = "results/GSI_and_Fst/ass-table-80-crop.pdf",
    ass_table80_tex = "tex/images/ass-table-80-crop.pdf",
    rosa_gsi_table_crop = "results/GSI_and_Fst/rosa-gsi-table-crop.pdf",
    rosa_gsi_table_tex = "tex/images/rosa-gsi-table-crop.pdf"
  )
}

@@

no_lfar_ass_table <- plot_self_assignment_table(top_ass$top, pop_labels, collection_order)

no_lfar_ass_table_80 <- top_ass$top %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

no_lfar_ass_table_Rosa <- top_ass$top %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

And then do the same for the data with the LFAR:
```{r}
lfar_sa <- self_assign(base_with_lfar, gen_start_col = 5)
lfar_top_ass <- get_top_assignments(lfar_sa, collection_order)
lfar_ass_table <- plot_self_assignment_table(lfar_top_ass$top, pop_labels, collection_order)

# make a version with no legend on it
lfar_ass_table_no_legend <- plot_self_assignment_table(lfar_top_ass$top, pop_labels, collection_order, no_legend = TRUE)

lfar_ass_table_80 <- lfar_top_ass$top %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

lfar_ass_table_Rosa <- lfar_top_ass$top %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

### Get some raw numbers

```{r}
# total number of fish
tot_fish <- nrow(lfar_top_ass$top)
tot_fish

# Get the number and fraction correct
lfar_top_ass$top %>%
  count(repunit == inferred_repunit) %>%
  mutate(fract = n/tot_fish)
```

```{r}
over80 <- nrow(lfar_top_ass$top %>% filter(repu_sclike > 0.8))

lfar_top_ass$top %>% filter(repu_sclike > 0.8) %>% count(repunit == inferred_repunit) %>% mutate(fract = n / over80)
```

Make a plot for > 80%
```{r}

# and then save that
ggsave(lfar_ass_table_80$plot, file = output_list$ass_table80, width = 10, height = 10.5)

# let's crop that down for use, too
CALL <- paste("pdfcrop ", output_list$ass_table80, collapse = " ")
system(CALL)

file.rename(from = output_list$ass_table80_crop, to = output_list$ass_table80_tex)
```

Make a plot for the byRosa:
```{r}

# and then save that
ggsave(lfar_ass_table_Rosa$plot, file = output_list$rosa_gsi_table, width = 10, height = 12.5)

# let's crop that down for use, too
CALL <- paste("pdfcrop ", output_list$rosa_gsi_table, collapse = " ")
system(CALL)

file.rename(from = output_list$rosa_gsi_table_crop, to = output_list$rosa_gsi_table_tex)
```

@@

We start by tallying up the misassigned fish---those that have a repunit and an
inferred repunit that differ:
```{r}
misassigned_ones <- lfar_top_ass$top %>%
  filter(repunit != inferred_repunit) %>%
  count(repunit, inferred_repunit)

@@

pick_em <- lapply(1:nrow(MO2), function(i) {

  FR <- MO2$Foc_repunit[i]
  FIR <- MO2$Foc_inferred_repunit[i]

  lfar_top_ass$repu_likes %>%
    ungroup() %>%
    filter(
      (repunit == FR & inferred_repunit == FIR) |
        (inferred_repunit == FR & repunit == FIR)
    ) %>%
    mutate(
      Foc_repunit = FR,
      Foc_inferred_repunit = FIR,
      .before = indiv
    ) 
}) %>%
    bind_rows()
```

OLD STUFF BELOW HERE
```{r}
tmp <- lapply(1:nrow(MO2), function(i) {
  lfar_top_ass$repu_likes %>%
    filter(
      (repunit == MO2$Foc_repunit[i] | repunit == MO2$Foc_inferred_repunit[i]) &
        (inferred_repunit == MO2$Foc_repunit[i] | inferred_repunit == MO2$Foc_inferred_repunit[i])
    ) %>%
    mutate(
      Foc_repunit =  MO2$Foc_repunit[i],
      Foc_inferred_repunit = MO2$Foc_inferred_repunit[i],
      .before = repunit
    )
}) %>%
  bind_rows() %>%
  arrange(indiv, Foc_repunit, Foc_inferred_repunit)

for_plots <- tmp %>%
  mutate(
    rep = ifelse(Foc_repunit == repunit, "rep", "inf"),
    inf = ifelse(Foc_inferred_repunit == inferred_repunit, "inf", "rep")
  ) %>%
  ungroup() %>%
  #count(Foc_repunit, Foc_inferred_repunit, repunit, inferred_repunit, rep, inf)
  select(-repunit, -inferred_repunit) %>%
  pivot_wider(names_from = c(rep, inf), names_sep = "_to_", values_from = repu_sclike) %>%
  mutate(condition = str_c(Foc_repunit, " misassigned to ", Foc_inferred_repunit), .after = Foc_inferred_repunit)
```

@@

I include the LFAR in their because I want to have it.  We note this is not for
demographic inference or anything.  It is just a summary of the differentiation
of the markers...

In order to do this we need to convert the data to a lumped format.
This means making integers of the alleles and then catenating them.
One locus has 11 allele, so we should use two digits for each allele.  We drop the
LFAR from this, because those were chosen specifically to be weird in the late fall.
```{r}
full_base_long <-  base_with_lfar %>% 
  select(-starts_with("NC_037130.1")) %>%  # drop the LFARs
  mutate(
    Pop = as.integer(factor(collection, levels = collection_order)),
    .before = collection
  ) %>%
  select(-repunit, -collection, -sample_type) %>%
  pivot_longer(
    cols = c(-indiv, -Pop),
    names_to = c("locus", "gene_copy"),
    names_pattern = "^(.*)_([12])$"
  )

@@

# long format with no missing data
long_genos <- base_with_lfar %>%
select(-repunit, -sample_type) %>%
pivot_longer(
cols = -c(indiv, collection),
names_to = c("locus", "gene_copy"),
values_to = "allele",
names_pattern = "(.*)[_.]([12])$"
) %>%
filter(!is.na(allele))

@@

# we need to get colors and things for the fill.  We just take the
# rubias output that was run through taf_prep and we keep the lower
# diagonal and we set all the cell_labels to NA.
lower_tri_base <- lfar_ass_table$RFL %>%
filter(as.integer(col_label) <= as.integer(row_label)) %>%
mutate(cell_label = NA_character_)

@@

# Get the RC_groups
RC_groups <- pop_labels %>%
    select(-old_name) %>%
    distinct() %>%
    rename(external_group = run_timing_group, internal_group = repunit, label = collection) %>%
    select(external_group, internal_group, label)

@@

```{r}
gsi_fst_fig_for_paper <- plot_grid(
  fst_with_scatters,
  plot_grid(
    get_legend(lfar_ass_table$for_external_legend),
    get_legend(lfar_ass_table$for_internal_legend),
    nrow = 1
  ),
  lfar_ass_table_no_legend$plot,
  nrow = 3,
  rel_heights = c(6,1.5,6)
)
